Simulation

### 2019-05-08 simulation.Rmd
Revised manuscript

```{r data-load}
dat_full <- read_csv("data/merged.csv", guess_max = 25000)
# Drop excluded subjects
dat <- filter(dat_full, !exclude)
# Drop unused factor levels
dat$subject <- fct_drop(dat$subject)

@@

This simulation examines potential bias in measures of metacognitive resolution (gamma correlation and $d_a$) due to guessing and task knowledge.

@@

```{r example-a}
example <- function(know=2, guesses=2, levels=6, n=10) {
  # Must sample letters because sample(x) automatically draws numbers <= x
  lls <- letters[1:levels]
  tibble(
    Know = c(rep(0, n-know), rep(1, know)),
    Accuracy = c(rep(0, (n-know-guesses)), rep(1, guesses+know)),
    Confidence = ordered(ifelse(Know==0, 
                                sample(lls[1:(levels/2)], n-know, T), 
                                sample(lls[(levels/2+1):levels], know, T)), 
                         levels = lls, labels = 1:levels)
  )[,-1]
}
set.seed(555)
a <- example(2, 2, 6, 10)
table(a)
vcdExtra::GKgamma(table(a))$gamma
summary(replicate(1000, vcdExtra::GKgamma(table(example(2, 2, 6, 10)))$gamma))
```

@@

We also verified a small variance between simulation runs due to large number of trials for each simulation.

```{r simsp-plot, layout="l-body-outset", fig.width=8, fig.height=6}
sims %>% 
  mutate(d_gen = ordered(d_gen)) %>% 
  filter(d_gen %in% c(0.55, 2.45)) %>% 
  mutate(Guessing = as.factor(p_guess)) %>% 
  gather(parameter, estimate, gamma, da) %>% 
  ggplot(aes(p_know, estimate, col = Guessing)) +
  scale_color_brewer(palette = "Set1", direction = -1) +
  geom_point(shape=1, alpha=.25) +
  stat_summary(fun.y=mean, geom="line") +
  stat_summary(fun.data=mean_sdl) +
  scale_x_continuous("Performance", breaks = pretty_breaks(5)) +
  scale_y_continuous("Estimated value", breaks = pretty_breaks(5)) +
  facet_grid(parameter~d_gen, scales = "free_y", labeller=label_both)
```

@@

We then tested whether the performance-resolution metric regression slopes were significantly steeper in the data, vs the equivalent slope in the simulations. We used second degree polynomial regressions. We tested

@@

```{r regression-test, layout="l-page", fig.width=8, fig.height=3.6, fig.cap="Comparison of observed correlations to simulation results."}
DEGREE <- 2
# PLot
tmp <- dat_metrics %>% 
  group_by(exp) %>% 
  summarise(d_ord = mean(d_ord, na.rm=T)) %>% 
  group_by(exp) %>% 
  mutate(d_gen = delta_bins[which.min(abs(d_ord - delta_bins))]) %>% 
  inner_join(sims_long) %>% 
  filter(p_guess == 0.25)
tmp %>% 
  ggplot(aes(p_know, estimate)) +
  scale_x_continuous("Knowledge", breaks = pretty_breaks(5)) +
  scale_y_continuous("Estimate", breaks = pretty_breaks(5)) +
  facet_grid(parameter~exp, scales="free_y") +
  theme(legend.position = "none") +
  geom_point(data = gather(dat_metrics, parameter, value, da, gamma), 
             aes(x = p_know, y = value), 
             col = "black", fill = "white", shape = 21, size = .5, alpha = .8) +
  geom_smooth(data = gather(dat_metrics, parameter, value, da, gamma), 
              aes(x = p_know, y = value), lty=2, 
              formula = y ~ poly(x, DEGREE),
              method = "lm", span=2, size=.66, se=T, alpha=.33, col="black") +
  geom_smooth(method = "lm", aes(col=d_gen), se=F, 
              formula = y ~ poly(x, DEGREE)) +
  scale_color_viridis_c("Resolution", 
                        values = rescale(tmp$d_gen, 
                                         to = c(0, 1))) +
  theme(panel.spacing = unit(11, "pt"))

tmp1 <- dat_metrics %>% 
  group_by(exp) %>% 
  summarise(d_ord = mean(d_ord, na.rm=T), s = mean(s, na.rm=T)) %>% 
  group_by(exp) %>% 
  mutate(d_gen = delta_bins[which.min(abs(d_ord - delta_bins))]) %>% 
  inner_join(sims_long_bias) %>% 
  ungroup() %>% 
  mutate(p_know = p_know-.5,
         p_know2 = p_know^2) %>% 
  nest(-exp, -parameter) %>% 
  mutate(
    res = map(data, ~lm(p_guess0.25 ~ p_know + p_know2, data = .)),
    out = map(res, ~tidy(.))
  ) %>% 
  unnest(out) %>% 
  select(exp, parameter, term, sim_coef = estimate)

# Test each experiment's slope against sim slopes
tmp2 <- dat_metrics %>% 
  gather(parameter, value, gamma:da) %>% 
  mutate(p_know = p_know-.5,
         p_know2 = p_know^2) %>% 
  nest(-exp, -parameter) %>% 
  mutate(
    res = map(data, ~lm(value ~ p_know + p_know2, data = .)),
    out = map(res, ~tidy(.)),
    N_subj = map_dbl(data, ~nrow(.))) %>% 
  unnest(out, .drop = T) %>%
  select(exp, N_subj, parameter, term, data_coef = estimate, std.error)

left_join(tmp1, tmp2) %>% 
  mutate(tval = (data_coef - sim_coef) / std.error,
         # One-sided p-values
         pval = pt(tval, df = N_subj-3, lower=FALSE)) %>% 
  filter(!(term %in% c("p_know2", "(Intercept)"))) %>%
  arrange(parameter) %>% 
  kable(
    digits = 2, 
    caption = "Tests of data regression slopes against simulations."
  )
```

### Debias average gammas

Next, we approximated what the average gamma & da in each experiment would have been, if guesses were absent. We did this by taking the average gamma, da, and performance of each experiment, and finding their corresponding non-guessing values from the simulations.

```{r}
tmp <- dat_metrics %>% 
  group_by(exp) %>% 
  summarise(
    p_know = mean(p, na.rm = T),
    gamma = mean(gamma, na.rm = T), 
    da = mean(da, na.rm = T)
    ) %>% 
  group_by(exp) %>% 
  # Get the nearest performance used in simulations
  mutate(p_know = p_know_bins[which.min(abs(p_know - p_know_bins))]) %>% 
  ungroup() %>% 
  gather(parameter, mean_data_estimate, gamma, da)
tmp2 <- sims_long_bias %>% 
  filter(between(p_know, .65, .76))
# Take rows for each exp and parameter where distance between observed resolution value and estimated(guessing=0.25) is the smalles
full_join(tmp, tmp2) %>% 
  mutate(difference = abs(mean_data_estimate - p_guess0.25)) %>% 
  arrange(parameter, exp, difference) %>%
  distinct(parameter, exp, .keep_all = TRUE) %>% 
  select(exp, p_know, parameter, mean_data_estimate, p_guess0.25, p_guess0) %>% 
  mutate(percent_increase = (p_guess0-mean_data_estimate)/mean_data_estimate) %>% 
  group_by(parameter) %>% 
  mutate(mpc = mean(percent_increase)) %>% 
  kable(
    escape = FALSE, 
    digits = 2,
    caption = "Resolution metrics from data"
  )
```

@@

Here, we attempt to test observed gamma/da against perfect metacognition. Each participant's observed gamma/da is paired with the corresponding mean simulated estimate at maximum d_gen (corresponding to about gamma = .96) and guessing = 0.25 and the participant's level of performance.

```{r}
tmp1 <- dat_metrics %>% 
  select(exp, subject, gamma, da, p_know) %>% 
  group_by(exp, subject) %>% 
  mutate(p_know_bin = p_know_bins[which.min(abs(p_know - p_know_bins))]) %>% 
  ungroup()
tmp2 <- sims_mean %>% 
  filter(d_gen == max(delta_bins), p_guess == "0.25")
tmp <- left_join(tmp1, tmp2, by = c("p_know_bin" = "p_know"))
names(tmp) <- str_replace_all(names(tmp), "\\.x", ".data")
names(tmp) <- str_replace_all(names(tmp), "\\.y", ".perfect")
# Are subjects gamma/da significantly lower than the gamma/da from simulation with perfect metacognition? That is, we use the difference between each subject's gamma and the simulated perfect gamma at that subjects point of performance
t.test(tmp$gamma.perfect, tmp$gamma.data, paired = T)
t.test(tmp$da.perfect, tmp$da.data, paired = T)
p1 <- tmp %>% 
  mutate(
    gamma_diff = gamma.perfect - gamma.data,
    da_diff = da.perfect - da.data
  ) %>% 
  select(exp, subject, contains("_diff")) %>% 
  gather(key, value, gamma_diff, da_diff) %>% 
  ggplot(aes(value)) +
  geom_histogram(bins = 50) +
  facet_wrap("key", scales = "free")
p2 <- tmp %>% 
  ggplot(aes(gamma.perfect, gamma.data)) +
  geom_abline(lty = 2) +
  geom_point()
p3 <- tmp %>% 
  ggplot(aes(da.perfect, da.data)) +
  geom_abline(lty = 2) +
  geom_point()
(p3 | p2) / p1
```

## Replication of Benjamin and Diaz

@@

```{r simsbd-plot1, layout="l-body-outset", fig.width=8, fig.height=6, fig.cap="Results from simulation replicating Benjamin and Diaz's (2008) simulation scheme when guessing=0. The two columns indicate different generative $\\sigma_s$s."}
sims_bd %>% 
  filter(p_guess == 0) %>% 
  gather(parameter, value, gamma, da, y0, da_ord, d_ord, r) %>% 
  group_by(parameter, d_gen, s_gen) %>% 
  summarise(n = n(), mean = mean(value, na.rm = T)) %>% 
  ggplot(aes(d_gen, mean, col = parameter)) +
  scale_color_brewer(
    "Parameter", 
    palette = "Dark2", 
    aesthetics = c("color", "fill")
    ) +
  geom_line(size=1.25) +
  geom_abline(lty=2, size = .33) +
  scale_x_continuous(Resolution, breaks = pretty_breaks(5)) +
  scale_y_continuous("Mean estimate", breaks = pretty_breaks(5)) +
  facet_wrap("s_gen", labeller=label_both) +
  theme(aspect.ratio = 1)
```

@@

```{r simsbd-plot-guessing, layout="l-body-outset", fig.width=8, fig.height=6, fig.cap="Results from simulation replicating Benjamin and Diaz's (2008) simulation scheme when guessing=0.25. The two columns indicate different generative $\\sigma_s$s."}
sims_bd %>% 
  filter(p_guess == 0.25) %>% 
  gather(parameter, value, gamma, da, y0, da_ord, d_ord, r) %>% 
  group_by(parameter, d_gen, s_gen) %>% 
  summarise(n = n(), mean = mean(value, na.rm = T)) %>% 
  ggplot(aes(d_gen, mean, col = parameter)) +
  scale_color_brewer(
    "Parameter", 
    palette = "Dark2", 
    aesthetics = c("color", "fill")
    ) +
  geom_line(size=1.25) +
  geom_abline(lty=2, size = .33) +
  scale_x_continuous(Resolution, breaks = pretty_breaks(5)) +
  scale_y_continuous("Mean estimate", breaks = pretty_breaks(5)) +
  facet_wrap("s_gen", labeller=label_both) +
  theme(aspect.ratio = 1)
```

@@

## Appendix A: Simulation function {.appendix}

@@

## Appendix B: Calculating gamma from ROC {.appendix}

@@

tmpp2 <- dat_metrics %>% 
  select(-r, -p, -con_m) %>% 
  mutate(az_zroc = pnorm(y0 / sqrt(1 + (1/s)^2 )),
         gamma_zroc = 2*az_zroc-1) %>%
  mutate(az_ord = pnorm( d_ord / sqrt(1 + (1/s_ord)^2 )),
         gamma_ord = 2*az_ord-1) %>% 
  ggplot(aes(gamma)) +
  geom_abline(lty=2) +
  theme(aspect.ratio=1)
tmpp2 + geom_point(aes(y=gamma_zroc)) | 
  tmpp2 + geom_point(aes(y=gamma_ord))

>>>>

```{r data-load}
dat <- read_csv("data/merged.csv", guess_max = 25000)
# Drop excluded subjects
# dat <- filter(dat_full, !exclude)
# Drop unused factor levels
dat$subject <- fct_drop(dat$subject)

@@

This simulation examines potential bias in measures of metacognitive resolution due to guessing and task knowledge.

@@

```{r example-a}
example <- function(know=2, guesses=2, levels=6, n=10) {
  # Must sample letters because sample(x) automatically draws numbers <= x
  lls <- letters[1:levels]
  tibble(
    Know = c(rep(0, n-know), rep(1, know)),
    Accuracy = c(rep(0, (n-know-guesses)), rep(1, guesses+know)),
    Confidence = ordered(
      ifelse(
        Know==0, 
        sample(lls[1:(levels/2)], n-know, T), 
        sample(lls[(levels/2+1):levels], know, T)
      ), 
      levels = lls, labels = 1:levels)
  )[,-1]
}
set.seed(555)
a <- example(2, 2, 6, 10)
table(a)
vcdExtra::GKgamma(table(a))$gamma
summary(replicate(1000, vcdExtra::GKgamma(table(example(2, 2, 6, 10)))$gamma))
```

@@

@@

We then tested whether the performance-resolution metric regression slopes were significantly steeper in the data, vs the equivalent slope in the simulations.

We tested

@@

```{r regression-test-fig, layout="l-page", fig.width=8, fig.height=3.6, fig.cap="Comparison of observed correlations to simulation results."}
DEGREE <- 1
# PLot
tmp <- dat_metrics %>% 
  group_by(exp) %>% 
  summarise(d_ord = mean(d_ord, na.rm=T)) %>% 
  group_by(exp) %>% 
  mutate(d_gen = delta_bins[which.min(abs(d_ord - delta_bins))]) %>% 
  inner_join(sims_long) %>% 
  filter(p_guess == 0.25)
tmp %>% 
  ggplot(aes(p_know, estimate)) +
  scale_x_continuous(
    "Knowledge", 
    breaks = pretty_breaks(5)
  ) +
  scale_y_continuous(
    "Estimate", 
    breaks = pretty_breaks(5)
  ) +
  theme(legend.position = "none") +
  geom_point(
    data = gather(mutate(dat_metrics, gamma=gz), parameter, value, da, gamma), 
    aes(x = p_know, y = value), 
    col = "black", fill = "white", shape = 21, size = .5, alpha = .8
  ) +
  geom_smooth(
    data = gather(mutate(dat_metrics, gamma=gz), parameter, value, da, gamma), 
    aes(x = p_know, y = value), lty=2, 
    formula = y ~ poly(x, DEGREE),
    method = "lm", size=.75, se=T, alpha=.33, col="black"
  ) +
  geom_smooth(
    method = "lm", aes(col=d_gen), se=F, 
    formula = y ~ poly(x, DEGREE)
  ) +
  scale_color_viridis_c(
    "Resolution", 
    values = rescale(tmp$d_gen, to = c(0, 1))
  ) +
  facet_grid(parameter~exp, scales="free_y") +
  theme(panel.spacing = unit(11, "pt"))
```

And then the tests

```{r regression-test}
# Simulated slope at each experiment's mean d_ord
tmp1 <- dat_metrics %>% 
  group_by(exp) %>% 
  summarise(d_ord = mean(d_ord, na.rm=T), s = mean(s, na.rm=T)) %>% 
  group_by(exp) %>% 
  mutate(d_gen = delta_bins[which.min(abs(d_ord - delta_bins))]) %>% 
  inner_join(sims_long_bias) %>% 
  ungroup() %>% 
  nest(-exp, -parameter) %>% 
  mutate(
    res = map(data, ~lm(p_guess0.25 ~ p_know, data = .)),
    out = map(res, ~tidy(.))
  ) %>% 
  unnest(out) %>% 
  select(exp, parameter, term, sim_coef = estimate)

# Observed slope for each experiment
tmp2 <- dat_metrics %>% 
  mutate(gamma = gz) %>% 
  gather(parameter, value, gamma:da) %>% 
  nest(-exp, -parameter) %>% 
  mutate(
    res = map(data, ~lm(value ~ p_know, data = .)),
    out = map(res, ~tidy(.)),
    N_subj = map_dbl(data, ~nrow(.))) %>% 
  unnest(out, .drop = T) %>%
  select(exp, N_subj, parameter, term, data_coef = estimate, std.error)

# Test each experiment's slope against sim slopes
left_join(tmp1, tmp2) %>% 
  mutate(
    tval = (data_coef - sim_coef) / std.error,
    # One-sided p-values
    pval = pt(tval, df = N_subj-2, lower=FALSE)
  ) %>% 
  filter(!(term %in% c("p_know2", "(Intercept)"))) %>%
  arrange(parameter) %>% 
  kable(
    digits = 2, 
    caption = "Tests of data regression slopes against simulations."
  )
```

@@

Here, we attempt to test observed gamma/da against near-perfect metacognition. Each participant's observed gamma/da is paired with the corresponding mean simulated estimate at maximum d_gen (corresponding to about gamma = .96) and guessing = 0.25 at the participant's level of performance.

```{r}
tmp1 <- dat_metrics %>% 
  select(exp, subject, gamma = gz, da, p_know) %>% 
  group_by(exp, subject) %>% 
  mutate(p_know_bin = p_know_bins[which.min(abs(p_know - p_know_bins))]) %>% 
  ungroup()
tmp2 <- sims_mean %>% 
  filter(d_gen == max(delta_bins), p_guess == "0.25")
tmp <- left_join(tmp1, tmp2, by = c("p_know_bin" = "p_know"))
names(tmp) <- str_replace_all(names(tmp), "\\.x", ".data")
names(tmp) <- str_replace_all(names(tmp), "\\.y", ".perfect")
mean(tmp$gamma.data)
mean(tmp$gamma.perfect)
# Are subjects gamma/da significantly lower than the gamma/da from simulation with perfect metacognition? That is, we use the difference between each subject's gamma and the simulated perfect gamma at that subjects point of performance
t.test(tmp$gamma.perfect, tmp$gamma.data, paired = T)
t.test(tmp$da.perfect, tmp$da.data, paired = T)
p1 <- tmp %>% 
  mutate(
    gamma_diff = gamma.perfect - gamma.data,
    da_diff = da.perfect - da.data
  ) %>% 
  select(exp, subject, contains("_diff")) %>% 
  gather(key, value, gamma_diff, da_diff) %>% 
  ggplot(aes(value)) +
  geom_histogram(bins = 50) +
  facet_wrap("key", scales = "free")
p2 <- tmp %>% 
  ggplot(aes(gamma.perfect, gamma.data)) +
  geom_abline(lty = 2) +
  geom_point()
p3 <- tmp %>% 
  ggplot(aes(da.perfect, da.data)) +
  geom_abline(lty = 2) +
  geom_point()
(p3 | p2) / p1
```

```{r}
# Difference from perfect on performance
# (How much below perfect are they as function of performance)
p4 <- tmp %>% 
  mutate(
    gamma_diff = gamma.perfect - gamma.data,
    da_diff = da.perfect - da.data
  ) %>% 
  select(exp, subject, p_know, contains("_diff")) %>% 
  gather(key, value, gamma_diff, da_diff)
p4 %>% 
  ggplot(aes(p_know, value)) +
  geom_point() +
  geom_smooth() +
  facet_wrap("key", scales = "free_y")
```

## Appendix A: Replication of Benjamin and Diaz

@@

```{r simsbd-plot1, layout="l-body-outset", fig.width=8, fig.height=6, fig.cap="Results from simulation replicating Benjamin and Diaz's (2008) simulation scheme when guessing=0. The two columns indicate different generative $\\sigma_s$s."}
sims_bd %>% 
  filter(p_guess == 0) %>% 
  gather(parameter, value, gamma, da, y0, da_ord, d_ord, r) %>% 
  group_by(parameter, d_gen, s_gen) %>% 
  summarise(n = n(), mean = mean(value, na.rm = T)) %>% 
  ggplot(aes(d_gen, mean, col = parameter)) +
  scale_color_brewer(
    "Parameter", 
    palette = "Dark2", 
    aesthetics = c("color", "fill")
    ) +
  geom_line(size=1.25) +
  geom_abline(lty=2, size = .33) +
  scale_x_continuous("Resolution", breaks = pretty_breaks(5)) +
  scale_y_continuous("Mean estimate", breaks = pretty_breaks(5)) +
  facet_wrap("s_gen", labeller=label_both) +
  theme(aspect.ratio = 1)
```

@@

```{r simsbd-plot-guessing, layout="l-body-outset", fig.width=8, fig.height=6, fig.cap="Results from simulation replicating Benjamin and Diaz's (2008) simulation scheme when guessing=0.25. The two columns indicate different generative $\\sigma_s$s."}
sims_bd %>% 
  filter(p_guess == 0.25) %>% 
  gather(parameter, value, gamma, da, y0, da_ord, d_ord, r) %>% 
  group_by(parameter, d_gen, s_gen) %>% 
  summarise(n = n(), mean = mean(value, na.rm = T)) %>% 
  ggplot(aes(d_gen, mean, col = parameter)) +
  scale_color_brewer(
    "Parameter", 
    palette = "Dark2", 
    aesthetics = c("color", "fill")
    ) +
  geom_line(size=1.25) +
  geom_abline(lty=2, size = .33) +
  scale_x_continuous("Resolution", breaks = pretty_breaks(5)) +
  scale_y_continuous("Mean estimate", breaks = pretty_breaks(5)) +
  facet_wrap("s_gen", labeller=label_both) +
  theme(aspect.ratio = 1)
```

@@

## Appendix B: Simulation function {.appendix}

@@

## Appendix C: Calculating gamma from ROC {.appendix}

@@

tmpp2 <- dat_metrics %>% 
  select(-r, -p, -con_m) %>% 
  mutate(az_zroc = pnorm(y0 / sqrt(1 + (1/s)^2 )),
         gamma_zroc = 2*az_zroc-1) %>%
  mutate(az_ord = pnorm( d_ord / sqrt(1 + (1/s_ord)^2 )),
         gamma_ord = 2*az_ord-1) %>% 
  ggplot(aes(gamma)) +
  geom_abline(lty=2) +
  theme(aspect.ratio=1)
(tmpp2 + geom_point(aes(y=gamma_zroc)) | 
  tmpp2 + geom_point(aes(y=gamma_ord))) &
  coord_cartesian(ylim=-1:1, xlim = -1:1)
